---
title: "Install"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
---

```{r, child="_setup.Rmd"}
```

The **DNAmArray**-package can be installed in several
ways which are described below. The package has been installed
successfully for >= R-3.2.0 on different linux-builds.

The package depends on many other packages from
[BioConductor](https://www.bioconductor.org) or
[cran](https://cran.r-project.org/). Usually these are installed
automatically, otherwise, we refer to
[BioConductor](https://www.bioconductor.org/install/) or
[cran](https://cran.r-project.org/doc/manuals/r-release/R-admin.html#Installing-packages)
documentation for the installation of these packages.

The installed package takes around ~7Mb of memory a large bit of it
comes from an `Rdata`-object containing the information on the
cross-hybridizing probes[@Chen2013].

## Install using the **devtools**-package ##

First install the package [**devtools**](https://github.com/hadley/devtools). Next use:

```{r devtools, eval=FALSE}
library(devtools)
install_git("https://git.lumc.nl/molepi/DNAmArray.git")
```

## Install from source using `git/R` ##

Using [git](https://git-scm.com/), e.g, use `git clone` and then build
and install the package from source:

```{r git, engine='bash', eval=FALSE}
git clone git@git.lumc.nl:molepi/DNAmArray.git
R CMD build DNAmArray
R CMD INSTALL DNAmArray_x.y.z.tar.gz
```
Change `_x.y.z.` to the proper version you downloaded!
