```{r, child="_setup.Rmd"}
```

***

# Functional normalization #

## Motivation ##

Our workflow uses the functional normalization approach<sup>16</sup>, which exploits internal control probes designed to detect technical variations without assaying biological differences. 

It has been shown to perform favourably when compared to other approaches<sup>17</sup>. Using the internal control probes avoids the problems associated with global normalization methods, where biological variation can be mistaken for a technical effect and removed. This is especially important in studies where groups are expected to have differential methylation signatures, such as multiple tissue studies<sup>18</sup>.

Conversations on the best approaches for normalization in DNAm data pipelines are ongoing<sup>19</sup>.

***

# Principal Components #

The default of selecting only two principal components is often too low for this type of data. Often you will see a drop-off in proportion of variance explained after a certain number of principal components, and this can indicate an efficient selection.

```{r 301scree}
var_explained %>% ggplot(aes(x=PC, y=var_explained)) +
  geom_line() +
  geom_point(color='grey5', fill='#6DACBC', shape=21, size=3) + 
  scale_x_continuous(breaks=1:ncol(pca$x)) +
  xlab("Principal Component") + 
  ylab("Proportion of variance explained") +
  theme_bw()
```

***

# Running Normalization #

In order to run normalization the annotation of the `RGset` must be updated for EPIC arrays.

```{r 302anno}
RGset@annotation <- c(array = "IlluminaHumanMethylationEPIC", annotation = "ilm10b4.hg19")
```

By default, functional normalization returns normalized copy number data making the returned `GenomicRatioSet` twice the size necessary when only beta-values or M-values are required. Therefore, we developed `preprocessFunnorm.DNAmArray()`, which adds an option not to return these to the `preprocessFunnorm()` function from [**minfi**](https://bioconductor.org/packages/release/bioc/html/minfi.html).

```{r 303funnorm, eval=F}
RGset <- preprocessFunnorm.DNAmArray(
  rgSet = RGset,
  nPCs = 4,
  keepCN = FALSE,
  verbose = TRUE
)
```

You can now extract the normalized beta values from this object. It can also be advisable to round the beta values to 4 significant figures, which both reduces the data size and provides a realistic measure of the available precision.

```{r 304getbetas}
betas <- getBeta(RGset, type="Illumina")
betas <- round(betas, 4)
```

```{r echo=F, warning=F, message=F}
betas <- normalizeQuantiles(betas)
```

If required, these can be transformed to M-values using [**minfi**](https://bioconductor.org/packages/release/bioc/html/minfi.html)'s `getM()` which applies a logit2-transformation to the beta-values.

***



