
***

# Introduction #

Our workflow uses the Functional Normalization approach (Hansen, 2014), which exploits internal control probes designed to detect technical variations without assaying biological differences. It has been shown to perform favourably when compared to other approaches (Heiss, 2015; Liu, 2016). 

Using the internal control probes avoids the problems associated with global normalization methods, where biological variation can be mistaken for a technical effect and removed. This is especially important in studies where groups are expected to have differential methylomes, such as multiple tissue studies (Min, 2018).

***

# Principal Components Selection #

The default of selecting only two principal components is often too low for this type of data, so our `screeplot()` function allows visualisation of control probe eigenvalues to help select the optimal number. Often you will see a drop-off in proportion of variance explained after a certain number of principal components, and this can indicate an efficient cut-off. 

The earlier screeplot can help us decide.

```{r}
var_explained %>% ggplot(aes(x=PC, y=var_explained)) +
  geom_line() +
  geom_point(color='grey5', fill='#6DACBC', shape=21, size=3) + 
  scale_x_continuous(breaks=1:ncol(pca$x)) +
  xlab("Principal Component") + 
  ylab("Proportion of variance explained") +
  theme_bw()
```

***

# Functional Normalization #

See Bas's email of 8th April 2025 on Normalization. Also Jazmin emailed about EPICv2 that day.

By default, functional normalization returns normalized copy number data making the returned `GenomicRatioSet` twice the size necessary when only beta-values or M-values are required. Therefore, we developed `preprocessFunnorm.DNAmArray()`, which adds an option not to return these to the `preprocessFunnorm()` function from [**minfi**](https://bioconductor.org/packages/release/bioc/html/minfi.html) (Feinberg, 2014).

Describe annotation update for all arrays including EPICv2

Get funnorm to work. Get MethyLIMP2 to work. split X Y auto

We have some missing values in the betas

```{r 302funnorm}
RGset@annotation <- c(array = "IlluminaHumanMethylationEPIC", annotation = "ilm10b4.hg19")
xtabs(~is.na(assay(RGset)))

RGset

library(impute) 

# Extract channels
red <- getRed(RGset)
green <- getGreen(RGset)

sum(is.na(red))
sum(is.na(green))

red_imputed <- impute.knn(red)
green_imputed <- impute.knn(green)

RGset_imputed <- RGset

assay(RGset_imputed, "Red") <- red_imputed$data
assay(RGset_imputed, "Green") <- green_imputed$data

xtabs(~is.na(assay(RGset_imputed)))


betas <- normalizeQuantiles(betas)
```

This function returns beta-values calculated with an offset of 100, similar to specifying `getBeta()` with option `type='Illumina'`. If required, these can be transformed to M-values using [**minfi**](https://bioconductor.org/packages/release/bioc/html/minfi.html)'s `getM()` which applies a logit2-transformation to the beta-values.

From inspecting the `GenomicRatioSet` object, you can see that there is one `Beta` assay, with methylation values for 485,512 CpG sites. 

***



