
```{r, child="_setup.Rmd"}
```

> **TODO extent descriptions** 

# Probe-level filtering #

Both the functions `probeFiltering` and `reduce` should be used to
obtain filtered beta- or M-values.

```{r probefiltering}
RGset <- probeFiltering(RGset)
```

## Number of beads ##

Occasionally some probes as measured with a too small number of beads
leading to unreliable probe intensities. By default probes intensities
are set to NA if the probe intensity is based on less then 3 beads. 

## Zero intensities ##

If there are no beads at all for a certain probe the probe intensity will
be exactly zero, these are set to NA, as well.

Samples which have more then 5% of their probes set to NA, based on
the above filter steps, will be removed from the `RGChannelSet`.

# CpG-level filtering #

## detection P-value ##

The negative controls on the 450k array can be used to detect outlying
samples e.g., samples that have a significant larger background then
average, this is summarized in the so-called detection
P-value.

Detection P-values are calculated using the `detectionP`-function from
[**minfi**](http:/bioconductor.org/packages/minfi). By default detection
P-values with a value >0.01 are set to NA.

## Cross reactive probes/CpGs ##

>THIS WILL BE REPLACE BY [@Zhou2017]

485,577 assays (482,421 CpG sites, 3091 non-CpG sites and 65 random
SNPs)...[@Bibikova2011]

Furthermore, several probes do not map uniquely to the genome or
contain common polymorphisms. Chen \emph{et al.}[@Chen2013] analyzed
in detail cross-reactive and polymorphic probes. We exclude the set of
polymorphic CG and CH probes (42.403) discovered by Chen.

482,421 CpG sites are measured using two chemistries Type I (135,476) and Type
II (350,036). the Type I chemistry uses two probes to measure a CpG site
while the Type II measure CpG methylation using one probe. The Type II
chemistry was specially designed to measure CpG methylation in low CpG
dense regions. There are 

Depending on the methylation status of the probe either of the green
or red fluorescence signals is detected a strongest.

Information on Type I and II probes, address, etc. can be extracted
using [**minfi**](http:/bioconductor.org/packages/minfi)-functionality.

```{r, eval=FALSE}
library(minfi)
locusNames <- getManifestInfo(RGset, "locusNames")
typeII <- getProbeInfo(RGset, type = "II")
nrow(typeII)
typeI <- getProbeInfo(RGset, type = "I")
head(typeI)
nrow(typeI)
```

# Success Rates #

All these probe filter-steps are implemented in the function
`filtering`.

RGset containing Red/Green signal intensities are converted to
Methylated and Unmethylated signals (see also, `?preprocessRaw`, from
minfi). These can in turn be converted to either beta-values or
M-values.


```{r reduce2beta, eval=FALSE}
betas <- reduce(GRset, RGset, what="beta")
```

```{r reduce2mvalues}
mvalues <- reduce(GRset, RGset, what="M")
```

```{r cleanup, echo = FALSE, results = 'hide'}
rm(GRset); gc()
```

# References #
