---
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
bibliography: DNAmArray.bib
---

```{r, child="_setup.Rmd"}
```

# Acknowledgements #

Paul Hop and Jenny van Dongen were involved in testing the first
versions and helped us improving the packages/workflow.

Contributors to the development of the package/workflow are: 

	Elmar W. Tobi,
	Roderick Slieker, 
	Wouter den Hollander, 
	Rene Luijk,
	Koen F.Dekkers and	
	Bas Heijmans (PI) from Molecular Epidemiology Section, Leiden
    University Medical Center, The Netherlands.

Development of the workflow and generation of example data in LLS was
funded by BBMRI-NL, a research infrastructure financed by the Dutch
government (NWO 184.021.007).

Maarten van Iterson is the main author of the package/workflow and
should be contacted for any questions: m.van_iterson at lumc.nl

Please cite Dekkers at al. [@dekkers2016] when using the 
**DNAmArray** package or workflow.

# disclamer #

**add one**

# licence # 

**add one**

```{r, child="_sessioninfo.Rmd"}
```
