
```{r, child="_setup.Rmd"}
```

***

# Acknowledgements #

Contributors to the development of the package and workflow are: 

*Elmar W. Tobi, Roderick Slieker, Wouter den Hollander, Rene Luijk, Koen F. Dekkers and Bas Heijmans (PI) from Molecular Epidemiology, Department of Biomedical Data Sciences, Leiden University Medical Center, Leiden, The Netherlands*

Development of the workflow and generation of previously used example data in LLS was funded by BBMRI-NL, a research infrastructure financed by the Dutch government (NWO 184.021.007).

Maarten van Iterson is the main author of the package and associated workflow. Lucy Sinke is the current author and maintainer and should be contacted with any questions: *l.j.sinke@lumc.nl*

Paul Hop and Jenny van Dongen were involved in testing and improving the pipeline.

Please cite van Iterson at al. [@VanIterson2017] when using the [**DNAmArray**](https://github.com/molepi/DNAmArray) package or workflow.

```{r, child="_sessioninfo.Rmd"}
```

# References #
