---
title: "Epigenomewide association study on Age"
output: 
  html_document:
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
---

```{r, child="_setup.Rmd"}
```

## Simple EWAS using **limma** ##

M-values valid to run EWAS using linear regression approach. 

Here skipping the empirical Bayes step, because of the large sample
size this should give very similar results?

```{r design1}
covariates <- c("DNA_BloodSampling_Age", "Sex", "Sentrix_Position", "Smoking", "Mono_Perc", "Baso_Perc", "Eos_Perc", "Neut_Perc") 
design <- data.frame(Int=1, colData(mvalues)[, covariates])
complete <- apply(design, 1, function(x) sum(is.na(x)) == 0 )
data <- assays(mvalues)$data
design <- design[complete,]
data <- data[,complete]
dim(data)
```

```{r design2}
design$Sentrix_Position <- as.integer(as.factor(design$Sentrix_Position))
```

```{r ewas}
library(limma)
fit <- lmFit(data, design)     
tstat <- fit$coef/fit$stdev.unscaled/fit$sigma        
pval <- 2*pnorm(-abs(tstat[,2]))
padj <- p.adjust(sort(pval, decreasing=FALSE), method="bonf")
head(padj[padj < 0.05])
```

## Unobserved covariates using **cate** ##

Often unobserved technical or biological covariates can be present. A
recent approach available from cran
[**cate**](https://cran.r-project.org/package=irlba) is able to
estimate the unobserved covariates. There are other methods as well,
...

```{r cate} 
library(cate)
fa <- cate.fit(design[,2,drop=FALSE], design[,-2], t(data), r = 3, calibrate=FALSE, adj.method="naive")
```

Here we assumed the presence of three unobserved covariates but cate
has two method implemented that can estimate the precise number of
unobserved covariates.

## Correct for bias and inflation ##

As is seen in GWAS, inflation, also occurs in EWAS. And not only
inflation but bias can be observed too. We developed a package called
[**bacon**](http://bioconductor.org/bacon) to estimate and correct for bias and inflation
for EWAS (and TWAS).

```{r bacon}
library(bacon)
tstats <- cbind(limma=tstat[,2], cate=as.vector(fa$beta.t))
bc <- bacon(tstats)
bc
print(plot(bc, type="hist"))
print(plot(bc, type="qq"))
head(pval(bc))
```


